Holomorphic field theories and higher algebra

### 2024-10-02 draft.tex
merge

\subsection{A loopy version}

@@

It is a remarkable fact, suggested by physicists, that there is also a map
\[
Y \colon {\rm Vac}[V] \to \End({\rm Vac}[V])[[z, z^{-1}]],
\]
known as the {\em vertex operator} or {\em state-field correspondence},
which is closely related to the $W[V]$-action.
The intuition behind $Y$ is that for any map $\phi: \CC^\times \to V$ and for any point $w \in \CC^\times$, 
there is a restriction of $\phi$ to a little disc around $w$.
That is, there is a restriction map $r_w: V((z)) \to V[[t]]$, where $t$ is the local coordinate $t = z-w$.
Hence there is a $w$-dependent map $\cO(V[[t]]) \to \cO(V((z)))$ by pulling back a function along $r_w$;
this map determines a $w$-dependent action of $\cO(V[[t]])$ on $\cO(V[[z]])$.
One can, in essence, quantize this map, by trying to extend the formulas of the $W[V]$-action,
and this leads to the map~$Y$.
If one axiomatizes the behavior of $Y$, one is led to the notion of a vertex algebra.

@@

At the level of action functionals, 
we can integrate $S_{2d}(\phi)$ over the angular coordinate to obtain the following formal sum of one-dimensional action functionals
\beqn
S_{1d}\left( \ldots, \phi_{-1}, \phi_0 , \phi_1, \ldots \right) = \sum_{k\in \ZZ} \int  \phi_{-k} \,\d \phi_{k} + k\, \phi_k \, \phi_{-k} \, \d t .
\eeqn 
\owen{The $H$ got dropped. Should we reintroduct it?}
Notice that the $k=0$ summand is just the Lagrangian of ordinary mechanics into $V$ with zero Hamiltonian. 
In general, we find mechanics into the symplectic vector space $LV$, where the symplectic structure uses the residue pairing in addition to the symplectic form on $V$, with a nontrivial Hamiltonian that encodes the $U(1)$-symmetry arising from rotating $LV$ in the natural way.

@@

There is a close relationship of the higher dimensional Weyl algebra to the higher dimensional $\beta\gamma$ system on $\CC^d - 0$ just as the loopy Weyl algebra is related to the $\beta\gamma$ system on~$\CC^\times$.

@@

\owen{discuss OPE, explain the shifted Poisson structure, and mention $\lambda$-brackets}

\subsubsection{From topological to holomorphic}

@@

In Section~\ref{???} we will examine another relationship --- generalizing the compactification/chiralization on $\CC$ --- that yields an $E_d$ algebra from a $\PDisc_d$-algebra.

@@

For a vertex algebra $\mathbb{V}$, the role of the circle is played by any elliptic curve,
and the role of Hochschild homology is played by conformal blocks on that genus one curve. 
Thus conformal blocks over elliptic curves provides the natural home for the characters of the vertex algebra.
The most basic example of the character of a conformal vertex algebra is its $q$-character,
which is defined in terms of a trace:
\beqn
\op{char}_{\mathbb{V}} (q) = \op{Tr}_{\mathbb{V}} (q^{L_0-c/24}) ,
\eeqn
where $L_0$ denotes the action of the vector field $z \del_z$.
The shift by the central charge guarantees that this expression has nice modularity properties, i.e., we can view $q$ as parametrizing the moduli space of elliptic curves.
\owen{Is that an ok turn of phrase?}

From the point of view of factorization algebras, we can understand this trace in terms of Hochschild homology of the \textit{modes} algebra associated to $\mathbb{V}$ as introduced in \S\ref{??}.
Indeed, suppose the conformal vertex algebra $\mathbb{V}$ arises from a holomorphic factorization algebra $\cF_{\mathbb{V}}$ on $\C$.
Then, if $r \colon \C^\times \to \RR_{>0}$ is the radial projection, we have discussed how to extract the modes algebra from the one-dimensional factorization algebra $r_* \cF_{\mathbb{F}}$.
If we further put this on the circle, we find Hochschild homology.
Thus, for every modulus $q$ one has
\beqn
\int_{E_q} \cF_{\mathbb{V}} = \int_{S^1} r_* \cF_{\mathbb{V}} = \op{Hoch}(r_* \cF_{\mathbb{V}}) .
\eeqn
\owen{I think $\int_{E_q}$ receives a map from $\op{Hoch}(r_* \cF_{\mathbb{V}})$ but I don't see why it's an isomorphism. On the other hand, given a conformal block (i.e., linear functional on $\int_{E_q}$, we can compose it with that map to define a trace on the modes algebra,
which ought to be the character.}
For more on this perspective see \cite{GuiLi}.
From this point of view, the factorization homology along an elliptic curve encodes characters in the sense of vertex algebras.

In higher dimensions we can build a similar picture,
replacing the moduli of elliptic curves with the moduli of {\it Hopf manifolds}.
Recall that a Hopf manifold is a quotient of the punctured affine plane
\beqn
H_{\mathbf{q}} = \left(\C^d - \{0\}\right) \slash \left((z_1,\ldots,z_d) \sim (q_1 z_1,\ldots,q_d z_d) \right)
\eeqn
where $\mathbf{q} = (q_1,\ldots,q_d)$ are complex numbers satisfying $0 < |q_i| < 1$.
For generic $\mathbf{q}$ such a complex manifold is diffeomorphic to $S^{2d-1} \times S^1$.
The factorization homology of a holomorphic factorization algebra $\cF$ along a Hopf surface of this kind can be identified with the Hochschild homology of the associative algebra of $S^{2n-1}$-modes:
\beqn
\int_{H_{\mathbf{q}}} \cF \simeq \op{Hoch}(r_* \cF)
\eeqn
where, again, $r\colon \C^d - \{0\} \to \RR_{>0}$ is the radius.
Thus, we see that the factorization homology $\int_{H_{\mathbf{q}}} \cF$ naturally encodes traces of the $S^{2n-1}$-mode algebra associated to $\cF$.
In \cite{SWchar}, it is shown how an important invariant called the \textit{supersymmetric index} (or supersymmetric partition function) of a supersymmetric theory along a sphere $S^{2n-1}$ be cast into this framework at the level the holomorphic twist.

>>>>

\subsection{A loopy version: the algebra of modes}

@@

It is a remarkable fact, suggested by physicists, that there is also a map
\[
Y \colon {\rm Vac}[V] \to \End({\rm Vac}[V])[[z, z^{-1}]],
\]
known as the {\em vertex operator} or {\em state-field correspondence},
which is closely related to the $W[V]$-action.
The intuition behind $Y$ is that for any map $\phi: \CC^\times \to V$ and for any point $w \in \CC^\times$, 
there is a restriction of $\phi$ to a little disc around $w$.
That is, there is a restriction map $r_w: V((z)) \to V[[t]]$, where $t$ is the local coordinate $t = z-w$.
Hence there is a $w$-dependent map $\cO(V[[t]]) \to \cO(V((z)))$ by pulling back a function along $r_w$;
this map determines a $w$-dependent action of $\cO(V[[t]])$ on $\cO(V[[z]])$.
One can, in essence, quantize this map, by trying to extend the formulas of the $W[V]$-action,
and this leads to the map~$Y$.
If one axiomatizes the behavior of $Y$, one is led to the notion of a vertex algebra.
The algebra $W[V]$ is known as the \textit{algebra of modes} associated to this vertex algebra.

@@

At the level of action functionals, 
we can integrate $S_{2d}(\phi)$ over the angular coordinate to obtain the following formal sum of one-dimensional action functionals
\beqn
S_{1d}\left( \ldots, \phi_{-1}, \phi_0 , \phi_1, \ldots \right) = \sum_{k\in \ZZ} \int  \phi_{-k} \,\d \phi_{k} + k\, \phi_k \, \phi_{-k} \, \d t .
\eeqn 
\owen{The $H$ got dropped. Should we reintroduct it?}
\brian{It's still there...}
Notice that the $k=0$ summand is just the Lagrangian of ordinary mechanics into $V$ with zero Hamiltonian. 
In general, we find mechanics into the symplectic vector space $LV$, where the symplectic structure uses the residue pairing in addition to the symplectic form on $V$, with a nontrivial Hamiltonian that encodes the $U(1)$-symmetry arising from rotating $LV$ in the natural way.

@@

There is a close relationship of the higher dimensional Weyl algebra to the higher dimensional $\beta\gamma$ system on $\CC^d - 0$ just as the loopy Weyl algebra is related to the $\beta\gamma$ system on~$\CC^\times$.
In this sense, $W_d[V]$ is an example of a `mode algebra', where the modes are now parametrized by the cohomology of punctured $d$-dimensional affine space.

@@

@@

In Section~\ref{???} we will examine another relationship --- generalizing the compactification/chiralization on $\CC$ --- that yields an $E_d$ algebra from a $\PDisc_d$-algebra.
\brian{This is probably a trivializable PDisc algebra by degree reasons?}

\brian{Maybe a better thing to say after this theorem is a `recognition' result.
That is, given a holomorphic PDisc$_{d}$ algebra, when is it equivalent to a algebra over $E_{2d}$ (in the essential image of the functor you wrote)?
Among possibly smaller things, this should require that the $\del_{\zbar}$'s are homotoped.}

@@

For a vertex algebra $\mathbb{V}$, the role of the circle is played by any elliptic curve,
and the role of Hochschild homology is played by conformal blocks on that genus one curve. 
Thus conformal blocks over elliptic curves provides the natural home for the characters of the vertex algebra.
The most basic example of the character of a conformal vertex algebra is its $q$-character,
which is defined in terms of a trace:
\beqn
\op{char}_{\mathbb{V}} (q) = \op{Tr}_{\mathbb{V}} (q^{L_0-c/24}) ,
\eeqn
where $L_0$ denotes the action of the vector field $z \del_z$.
The shift by the central charge guarantees that this expression has nice modularity properties, i.e., we can view $q$ as parametrizing the moduli of elliptic curves.

From the point of view of factorization algebras, we can understand this trace in terms of Hochschild homology of the algebra of modes associated to $\mathbb{V}$.
Indeed, suppose the conformal vertex algebra $\mathbb{V}$ arises from a holomorphic $S^1$-equivariant factorization algebra $\cF_{\mathbb{V}}$ on $\C$.
If $r \colon \C^\times \to \RR_{>0}$ is the radial projection there is a procedure for extracting the associative algebra of modes from the one-dimensional factorization algebra $r_* \cF_{\mathbb{V}}$.
Indeed, while $r_* \cF_{\mathbb{V}}$ may not be a locally constant factorization algebra, it always contains a dense subfactorization algebra which is locally constant.\footnote{To extract it, one assumes that $\cF$ is $S^1$-equivariant.
For each interval $I \subset \RR_{>0}$ one then looks at the subspace of $\cF(\pi^{-1}(I))$ consisting of the eigenspaces for this $S^1$ action.
The assignment of an interval $I$ to this subspace is a locally constant factorization algebra.}
With slight abuse of notation, we will still denote $r_* \cF_{\mathbb{F}}$ the locally constant factorization algebra obtained from this compactification.

If we further put this on the circle, we find Hochschild homology.
Thus, for each modulus~$q$ functoriality of factorization homology gives rise to a map
\beqn
\int_{S^1_{\sigma}} r_* \cF_{\mathbb{V}} = \op{Hoch}(r_* \cF_{\mathbb{V}}, \sigma) \to \int_{E_q} \cF_{\mathbb{V}} .
\eeqn
Here, $\sigma$ is an automorphism defined in terms of the given $S^1$-action on $\cF_{\mathbb{V}}$, and what appears on the left-hand side is the $\sigma$-twisted Hochschild homology.
For free theories, this map is a quasi-isomorphism.
On the other hand, given a conformal block (i.e., linear functional on $\int_{E_q} \cF_{\mathbb{V}}$) we can compose it with that map to define a trace on the modes algebra.
From this point of view, the factorization homology along an elliptic curve encodes characters in the sense of vertex algebras.
For more on this perspective see \cite{GuiLi}, where they carefully develop the notion of a trace for chiral/vertex algebras.

In higher dimensions we can build a similar picture,
replacing the moduli of elliptic curves with the moduli of {\it Hopf manifolds}.
Recall that a Hopf manifold is a quotient of the punctured affine plane
\beqn
H_{\mathbf{q}} = \left(\C^d - \{0\}\right) \slash \left((z_1,\ldots,z_d) \sim (q_1 z_1,\ldots,q_d z_d) \right)
\eeqn
where $\mathbf{q} = (q_1,\ldots,q_d)$ are complex numbers satisfying $0 < |q_i| < 1$.
For generic $\mathbf{q}$ such a complex manifold is diffeomorphic to $S^{2d-1} \times S^1$.
The factorization homology of a holomorphic factorization algebra $\cF$ along a Hopf surface of this kind can be identified with the Hochschild homology of the associative algebra of $S^{2n-1}$-modes:
\beqn
\op{Hoch}(r_* \cF , \sigma) \to \int_{H_{\mathbf{q}}} \cF 
\eeqn
where, again, $r\colon \C^d - \{0\} \to \RR_{>0}$ is the radius and $\sigma$ is an automorphism of $\cF$ induced from the $S^1$-equivariant structure of $\cF$.
Thus, we see that the factorization homology $\int_{H_{\mathbf{q}}} \cF$ naturally encodes traces of the $S^{2n-1}$-mode algebra associated to $\cF$.
In \cite{SWchar}, it is shown how an important invariant called the \textit{supersymmetric index} (or supersymmetric partition function) of a supersymmetric theory along a sphere $S^{2n-1}$ be cast into this framework at the level the holomorphic twist.
